# Managing R and R packages {#Rpackages}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE) 
options(rt.theme = "lightgrid")
options(rt.fit.theme = "lightgrid")
```

```{r, comment="", results="asis", echo=FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE)
```

## R
This book was compiled using `r R.Version()$version.string`.  
Make sure you have the latest version by visiting the [R project website](https://www.r-project.org)

## R packages
### CRAN
The Comprehensive R Archive Network is the official R package repository and currently hosts 15338 packages. To install a package from CRAN, use the builtin `install.packages` command:
```{r eval=FALSE}
install.packages('glmnet')
```

#### Check for outdated packages
```{r eval = FALSE}
old.packages()
```

#### Update installed packages
If you don't set `ask = FALSE`, you will have to accept each package update separately.
```{r eval = FALSE}
update.packages(ask = FALSE)
```

### GitHub
GitHub contains a large number of R packages, some of which also exist in CRAN, but the GitHub version may be updated a lot more frequently. To install from GitHub, you need to have the `remotes` package from CRAN first:
```{r eval = FALSE}
install.packages("remotes")
```

```{r eval = FALSE}
remotes::install_github("username/reponame")
```

Note: Running `remotes::install_github("user/repo")` will not reinstall a previously installed package, unless it has been updated.

### Bioconductor
Bioconductor is a repository which includes tools for the analysis and comprehension of high-throughput genomic data, among others. To install package from Bioconductor, first install the `BiocManager` package from CRAN:
```{r eval = FALSE}
install.packages(“BiocManager")
```
and then use that similar to the builtin `install.packages`:
```{r eval = FALSE}
BiocManager::install("packageName")
```

### Installed packages
List all R packages installed on your system with `installed.packages()` (the following block has not been run to prevent a very long output)
```{r eval = FALSE}
installed.packages()
```

List attached packages with `search()`:
```{r}
search()
```

List attached packages with their system path:
```{r}
searchpaths()
```




## RStudio IDE
[RStudio](https://rstudio.com/) is an Integrated Development Environment ([IDE](https://en.wikipedia.org/wiki/Integrated_development_environment)) for R, which can make any work in R easier, more productive, and more fun. Make sure to keep your installation up-to-date as new features 