# Preface {-}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE) 
options(rt.theme = "lightgrid")
options(rt.fit.theme = "lightgrid")
```

```{r, comment="", results="asis", echo=FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE)
```

Throughout this book you will see boxes with R code followed by its output, if any. The code (or input) is decorated with a teal border on the left to separate it from its output, like in the following example:

```{r}
x <- rnorm(200)
x[1:20]
```

Notice that R adds numbers in brackets in the beginning of each row. This happens when R prints the contents of a vector. The number is the integer index of the first element in that row. Therefore, the first one is always `[1]` and the number of the subsequent rows depends on how many elements fit in each line. If the output is a single element, it will still have `[1]` in front of it.  

Also notice that if we enclose the assignment operation of a variable in parentheses, this prints the resulting value of the variable. Therefore, this:

```{r}
(y <- 4)
```

is equivalent to:

```{r}
y <- 4
y
```

Note that if you mouse over the input code box, a clickable "Copy to clipboard" appears on the top right of the box allowing you to copy paste into an R session or file.

Lastly, you will see the following Info, Note, or Warning boxes at times:

```{block, type="Info"}
This is some info
```

```{block, type="Note"}
This is a note
```

```{block, type="Warning"}
This is a warning
```