# R setup, Packages & Docs {#Rsetup}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE)
```

## R

This book was compiled using `r R.Version()$version.string`.  
Make sure you have the latest version by visiting the [R project website](https://www.r-project.org)

It's a good idea to keep a log of the version of R and installed packages when beginning a new project. An easy way to do this is to save the output of `sessionInfo()`:

```{r}
sessionInfo()
```

## R packages
### CRAN
The Comprehensive R Archive Network (CRAN) is the official R package repository and currently hosts 16271 packages (as of 2020-09-13). To install a package from CRAN, use the builtin `install.packages` command:
```{r eval=FALSE}
install.packages('glmnet')
```

#### Check for outdated packages

```{r eval = FALSE}
old.packages()
```

#### Update installed packages

If you don't set `ask = FALSE`, you will have to accept each package update separately.

```{r eval = FALSE}
update.packages(ask = FALSE)
```

### GitHub

GitHub contains a large number of R packages, some of which also exist in CRAN, but the GitHub version may be updated a lot more frequently. To install from GitHub, you need to have the `remotes` package from CRAN first:

```{r eval = FALSE}
install.packages("remotes")
```

```{r eval = FALSE}
remotes::install_github("username/reponame")
```

Note: Running `remotes::install_github("user/repo")` will not reinstall a previously installed package, unless it has been updated.

### Bioconductor

Bioconductor is a repository which includes tools for the analysis and comprehension of high-throughput genomic data, among others. To install package from Bioconductor, first install the `BiocManager` package from CRAN:

```{r eval = FALSE}
install.packages(“BiocManager")
```

and then use that similar to the builtin `install.packages`:

```{r eval = FALSE}
BiocManager::install("packageName")
```

### Installed packages

List all R packages installed on your system with `installed.packages()` (the following block has not been run to prevent a very long output)

```{r eval = FALSE}
installed.packages()
```

List attached packages with `search()`:

```{r}
search()
```

List attached packages with their system path:

```{r}
searchpaths()
```


## RStudio IDE

[RStudio](https://rstudio.com/) is an Integrated Development Environment ([IDE](https://en.wikipedia.org/wiki/Integrated_development_environment)) for R, which can make work in R easier, more productive, and more fun. Make sure to keep your installation up-to-date; new features are added often.  

It is recommended to set up a new RStudio project for each data project: Select File > New Project... from the main menu.


## Builtin Documentation

After you've successfully installed R and RStudio, one of the first things to know is how to access and search the builtin documentation.  

### Get help on a specific item

If you know the name of what you're looking for (an R function most commonly, but possibly also the name of a dataset, or a package itself), just type `?` followed by the name of said function, dataset, etc. in the R prompt:

```{r}
?sample
```

In RStudio, the above example will bring up the documentation for the `sample` function in the dedicated "Help" window, commonly situated at the bottom right (but can be moved by the user freely). If you are running R directly at the system shell, the same information is printed directly at the console.  
Try running the above example on your system.

### Search the docs

If you do not know the name of what you are looking for, you can use double question marks, `??`, followed by your query (this is short for the `help.search` command that provides a number of arguments you can look up using `?help.search`):

```{r}
??bootstrap
```
