# Basic operations {#basicops}

First, before even learning about data types and structures, it may be worth looking at some of the basic mathematical and statistical operations in R.

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE) 
```

## Arithmetic

```{r}
x <- 10
y <- 3
```

Standard arithmetic operations are as expected:

```{r}
x + y
x - y
x * y
x / 3
```
Exponentiation uses `^`:
(This is worth pointing out, because while `^` is likely the most common way to represent exponentiation, the symbol used for exponentiation varies across languages)

```{r}
x^3
```

Square root is `sqrt()`:

```{r}
sqrt(81)
```

Integer division 
i.e. *Divide and forget the remainder*

```{r}
x %/% 3
```

i.e. how many times the denominator fits in the numerator, without taking fractions of the denominator. It can be applied on decimals the same way:

```{r}
9.5 %/% 3.1
```

Modulo operation
i.e. *Divide and return just the remainder*

```{r}
x %% y
```
```{r}
x <- (-10:10)[-11]
y <- sample((-10:10)[-11], 20)
x - (x %/% y) * y == x %% y
```
Try to figure out what the following does:  

```{r}
x <- rnorm(20)
y <- rnorm(20)
round(x - (x %/% y) * y, 5) == round(x %% y, 5)
```

```{block, type="info"}
`round(x, digits)` rounds `x` to the desired number of digits and is used to overcome rounding errors.
```
## Logical operations

Logical AND with `&`

```{r}
T & T
```

```{r}
T & F
```

Logical OR with `|`

```{r}
T | F
```

Logical negation with `!`

```{r}
x <- TRUE
!x
```

Exclusive OR with `xor()`
(= one or the other is TRUE but not both)

```{r}
a <- c(T, T, T, F, F, F)
b <- c(F, F, T, F, T, T)
a & b
a | b
xor(a, b)
```

Test all elements are TRUE with `all()`:

```{r}
all(a)
```

Test if any element is TRUE with `any()`:

```{r}
any(a)
```

## Common descriptive stats

Let's use the `rnorm` function to draw 200 numbers from a random normal distribution:

```{r}
x <- rnorm(200)
```

Basic descriptive stat operations:

```{r}
mean(x)
median(x)
sd(x) # standard deviation
min(x)
max(x)
range(x)
```
```{r}
summary(x)
```
