# Math & Stat Ops {#mathops}

First, before even learning about data types and structures, it may be worth looking at some of the basic mathematical and statistical operations in R.

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE) 
```

## Arithmetic

```{r}
x <- 10
y <- 3
```

Standard arithmetic operations are as expected:

```{r}
x + y
x - y
x * y
x / 3
```
Exponentiation uses `^`:
```{r}
x^3
```

Integer division 
i.e. *Divide and forget the remainder*
```{r}
x %/% 3
```

i.e. how many times the denominator fits in the numerator, without taking fractions of the denominator. It can be applied on decimals the same way:

```{r}
9.5 %/% 3.1
```

Modulo operation
i.e. *Divide and return just the remainder*
```{r}
x %% y
```
## Basic stat operations

Let's use the `rnorm` function to draw 200 numbers from a random normal distribution:
```{r}
x <- rnorm(200)
```

Basic descriptive stat operations:
```{r}
mean(x)
median(x)
sd(x) # standard deviation
min(x)
max(x)
```

