# Data Types {#types}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE) 
options(rt.theme = "lightgrid")
options(rt.fit.theme = "lightgrid")
```

```{r, comment="", results="asis", echo=FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE)
```

## Base types

```{block, type="rmdtip"}
All data in R is stored in vectors, whether stand-alone, i.e. a 1-D collection of items of the same 
type (e.g. numeric, character, etc.), or within another [data structure](#structures) (e.g. a data.frame,
list).

**Data types in R are essentially different types of vectors.**
```

R includes a number of builtin data types.  

These are defined by the R core team: users cannot define their own data types.  

Users can define their own classes - see [Classes and Object-Oriented Programming](#classes).  

Some of the most popular data types in R are:

* Logical (i.e. TRUE or FALSE, a.k.a. [Boolean](https://en.wikipedia.org/wiki/Boolean_data_type))
* Numeric, integer
* Numeric, [double](https://en.wikipedia.org/wiki/Double-precision_floating-point_format)
* Character
* Environment
* Closure (essentially, function; technically, a function and an environment)

```{block, type="rmdcaution"}
Many errors in R occur because a variable is, or gets coerced to, the wrong type by accident.
```

```{block, type="rmdtip"}
**Check variable types with `typeof()` and/or `str()`.**
```

## Assignment

Use `<-` for all assignments:

```{r}
x <- 3
# You can add comments within code blocks using the usual "#" prefix
```

```{block, type="rmdtip"}
In [RStudio](https://rstudio.com/) the keyboard shortcut for the assignment operator `<-` is `Option -` (MacOS) or `Alt -` (Windows).
```

Typing the name of an object...

```{r}
x
```
...is equivalent to printing it

```{r}
print(x)
```

You can also place any assignment in parentheses and this will perform the assignment and print the object:

```{r}
(x <- 3)
```

```{block, type="rmdnote"}
While you *could* use the equal sign '=' for assignment, you should only use it to pass arguments to functions.
```

You can assign the same value to multiple objects - this can be useful when initializing variables.

```{r}
x <- z <- init <- 0
x
z
init
```

Excitingly, R allows assignment in the opposite direction as well:

```{r}
10 -> x
x
```

We shall see later that the `->` assignment can be convenient at the end of a [pipe](https://class.lambdamd.org/progdatscir/functions.html#the-pipe-operator).

You can even do this, which is fun (?) but unlikely to be useful:

```{r}
x <- 7 -> z
x
z
```

## Combine values

Use `c()` to combine multiple values into a vector:

```{r}
x <- c(-12, 3.5, 104)
x
```


## Initialize - coerce - test (types)

The following summary table lists the functions to *initialize*, *coerce* (=convert), and *test* the core data types, which are shown in more detail in the following paragraphs:

| **Initialize** | **Coerce**        | **Test**          |
|---------------:|------------------:|------------------:|
| `logical(n)`   | `as.logical(x)`   | `is.logical(x)`   |
| `integer(n)`   | `as.integer(x)`   | `is.integer(x)`   |
| `double(n)`    | `as.double(x)`    | `is.double(x)`    |
| `numeric(n)`   | `as.numeric(x)`   | `is.numeric(x)`   |
| `character(n)` | `as.character(x)` | `is.character(x)` |

Note: `numeric` and `double` functions on lines 3 and 4 above are equivalent. 
(Try printing `numeric` and `double` in the console)

## Logical 

If you are writing code, (it's good practice to) use `TRUE` and `FALSE`.  
On the console, you can abbreviate to `T` and `F`, they are the same.

```{r}
a <- c(TRUE, FALSE)
a <- c(T, F)
x <- 4
b <- x > 10
b
str(b)
typeof(b)
```

## Integer

Create a range of integers using colon notation `start:end`:

```{r}
(x <- 11:15)
typeof(x)
str(x)
```

Note that assigning an integer defaults to type double:

```{r}
x <- 1
typeof(x)
str(x)
```

You can force it to be stored as integer by adding an `L` suffix:

```{r}
x <- 1L
typeof(x)
str(x)
```
```{r}
x <- c(1L, 3L, 5L)
str(x)
```

## Double

```{r}
x <- c(1.2, 3.4, 10.987632419834556)
x
typeof(x)
str(x)
```

## Character

A character vector consists of one or more elements, each of which consists of one or more actual characters, i.e. it is **not** a vector of single characters. (The length of a character vector is the number of individual elements, and is not related to the number of characters in each element)

```{r}
x <- "word"
typeof(x)
length(x)
```

```{r}
(x <- c("a", "b", "gamma", "delta"))
typeof(x)
length(x)
```

## Environment

Defining your own environments is probably for advanced use only:

```{r}
x <- new.env()
x$name <- "Guava"
x$founded <- 2020
x
typeof(x)
```

## Closure (function)

Closures are functions - they contain their own variable definitions.  
Read more on [functions](#functions).

```{r}
square <- function(x) x^2
square(3)
typeof(square)
```

## Initialize vectors

You can create / initialize vectors of specific type with the `vector` command and specifying a `mode` or directly by calling the relevant function:

```{r}
(xl <- vector(mode = "logical", length = 10))
(xd <- vector(mode = "double", length = 10))
(xn <- vector(mode = "numeric", length = 10)) # same as "double"
(xi <- vector(mode = "integer", length = 10))
(xc <- vector(mode = "character", length = 10))
```

These are aliases of the `vector` command above (print their source code to see for yourself)

```{r}
xl <- logical(10)
xd <- double(10)
xn <- numeric(10) # same as double
xi <- integer(10)
xc <- character(10)
```

## Explicit coercion

We can explicitly convert objects of one type to a different type using `as.*` functions:

```{r}
(x <- c(1.2, 2.3, 3.4))
(x <- as.logical(x))
(x <- as.double(x))
(x <- as.numeric(x))
(x <- as.integer(x))
(x <- as.character(x))
```

Logical vectors are converted to 1s and 0s as expected:  

TRUE becomes 1 and FALSE becomes 0

```{r}
x <- c(TRUE, TRUE, FALSE)
as.numeric(x)
```

Note that converting from numeric to logical **anything other than zero is TRUE**:

```{r}
x <- seq(-2, 2, .5)
as.logical(x)
```

Not all conversions are possible.  
There is no meaningful/consistent way to convert a character vector to numeric.  
The following outputs NA values and prints a (helpful) error message.

```{r}
x <- c("mango", "banana", "tangerine")
as.numeric(x)
```

## Implicit coercion

Remember, the language generally tries to make life easier. Sometimes this means it will automatically coerce one class to another to allow requested operations.

For example, you can get the sum of a logical vector.  
It will automatically be converted to numeric as we saw earlier.

```{r}
x <- c(TRUE, TRUE, FALSE)
sum(x)
```

On the other hand, you cannot sum a [factor](#factors), for example.  
You get an error with an explanation:

```{r error = TRUE}
x <- factor(c("mango", "banana", "mango"))
sum(x)
```

```{block, type="rmdnote"}
Note: We had to add `error = TRUE` in the Rmarkdown's code block's options (not visible in the HTML output), because otherwise compilation of the Rmarkdown document would stop at the error.
```

If for some reason it made sense, you could explicitly coerce to numeric and then sum:

```{r}
x <- factor(c("mango", "banana", "mango"))
sum(as.numeric(x))
```

## `NA`: Missing Values

Missing values in any data type - logical, integer, double, or character - are coded using `NA`.  
To check for the presence of `NA` values, use `is.na()`:

```{r}
(x <- c(1.2, 5.3, 4.8, NA, 9.6))
is.na(x)
```

```{r}
(x <- c("mango", "banana", NA, "sugar", "ackee"))
is.na(x)
```

```{r}
(x <- c(T, T, F, T, F, F, NA))
is.na(x)
```

`is.na()` works similarly on matrices:

```{r}
x <- matrix(1:20, 5)
x[4, 3] <- NA
is.na(x)
```

```{block, type="rmdnote"}
Note that `is.na()` returns a response for each element (i.e. is [vectorized](#vectorization)) in contrast to `is.numeric()`, `is.logical()`, etc. It makes sense, since the latter are chacking the type of a whole object, while the former is checking individual elements. 
```

`anyNA()` is a very useful function to check if there an any NA values in an object:

```{r}
anyNA(x)
```

```{block, type="rmdnote"}
Any operations on an `NA` results in `NA`
```

```{r}
x <- c(1.2, 5.3, 4.8, NA, 9.6)
x*2
```
Multiple functions that accept as input an object with multiple values (a vector, a matrix, a data.frame, etc.) will return `NA` if *any* element is `NA`:

```{r}
mean(x)
median(x)
sd(x)
min(x)
max(x)
range(x)
```

First, make sure `NA` values represent legitimate missing data and not some error.  
Then, decide how you want to handle it.

In all of the above commands you can pass `na.rm = TRUE` to ignore `NA` values:

```{r}
mean(x, na.rm = TRUE)
median(x, na.rm = TRUE)
sd(x, na.rm = TRUE)
min(x, na.rm = TRUE)
max(x, na.rm = TRUE)
range(x, na.rm = TRUE)
```

More generally, you can use `na.exclude()` to exclude NA values from R objects. This can be very useful for function that do not include a `na.rm` or similar argument to handle `NA` values.

```{r}
x <- c(1, 2, NA, 4)
na.exclude(x)
```

On a data.frame, `na.exclude()` excludes rows with any `NA`s:

```{r}
df <- data.frame(a = c(1, 2, NA, 4),
                 b = c(11, NA, 13, 14))
na.exclude(df)
```

The chapter on [Handling Missing Data](#missingdata) describes some approaches to handling missing data in the context of statistics or modeling, commonly supervised learning.

## `NaN`: Not a number

`NaN` is a special case of `NA` and can be the result of undefined mathematical operations:

```{r}
a <- log(-4)
```

Note that `class()` returns "numeric":

```{r}
class(a)
```

To test for `NaN`s, use:

```{r}
is.nan(a)
```
`NaN`s are also `NA`:

```{r}
is.na(a)
```
But the opposite is not true:

```{r}
is.nan(NA)
```
```{block, type="rmdnote"}
`NaN` can be considered a subtype of `NA`, as such: `is.na(NaN)` is `TRUE`, but `is.nan(NA)` is `FALSE`.
```


## `NULL`: the empty object

The `NULL` object represents an empty object.

```{block, type="rmdnote"}
`NULL` means empty, ***not missing***, and is therefore entirely different from `NA`
```

`NULL` shows up for example when initializing a list:

```{r}
a <- vector("list", 4)
a
```

and it can be replaced normally:

```{r}
a[[1]] <- 3
a
```

### Replacing with NULL

You cannot replace one or more elements of a vector/matrix/array with `NULL` because `NULL` has length 0 and replacement requires object of equal length:

```{r, error=T}
a <- 11:15
a
a[1] <- NULL
```

However, in lists and therefore also data frames, replacing an element with `NULL` ***removes that element***:

```{r}
al <- list(alpha = 11:15,
           beta = rnorm(10),
           gamma = c("mango", "banana", "tangerine"))
al
al[[2]] <- NULL
al
```

Finally, `NULL` is often used as the default value in a function's argument. The function definition must then determine what the default behavior/value should be.
