# Indexing - Subsetting - Slicing {#indexing}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE) 
```

An index is used to select elements of a vector, matrix, array, list or data frame.  
You can select (or exclude) one or multiple elements at a time.  

An index is one of two types:

- logical index: for each elements in an object specify `TRUE` if you want to include it, or `FALSE` to exclude it from the selection.
- integer index: define the position of elements to select.  

The main indexing operator in R is the square bracket `[`.
Indexing in R is 1-based, meaning the first item of a vector is in position 1.  
(If you are wondering why we even have to mention this, know that many programming languages use [0-based indexing](https://en.wikipedia.org/wiki/Zero-based_numbering#Computer_programming))

## Vectors {#indexvectors}

```{r}
x <- rnorm(10)
x
```

Get the 5th element of a vector:

```{r}
x[5]
```

Get elements 6 through 9 of the same vector:

```{r}
x[6:9]
```

## Matrices {#indexmatrices}

Reminder:

- A matrix is a 2D vector and contains elements of one type only (numeric, integer, character, factor).  
- A data frame is a 2D list and each column can contain different type of data.

Numeric matrix:

```{r}
mat <- matrix(1:40, 10)
df <- as.data.frame(mat)
df
```

To index a 2D structure, whether a matrix or data frame, we use `[row, column]`

To get the contents of the fifth row, second column:

```{r}
mat[5, 2]
df[5, 2]
```

If you want to select an entire row or an entire column, you leave the row or column index blank, but - necessarily - use a comma:

Get the first row:

```{r}
df[1, ]
```

Get the second column:

```{r}
df[, 2]
```

### Range of rows and columns

You can define ranges for both rows and columns:

```{r}
df[6:7, 2:4]
```

You can return rows and/or columns reversed too if desired:

```{r}
df[7:6, 4:2]
```

Or use vectors to specify of any rows and columns:

```{r}
df[c(2, 4, 7), c(1, 4, 3)]
```

### Matrix of indexes

This is quite less common, but potentially useful. It allows you to specify a series of individual `[i, j]` indexes (i.e. is a way to select multiple non-contiguous elements)

```{r}
idm <- matrix(c(2, 4, 7, 4, 3, 1), 3)
idm
```

An n-by-2 matrix can be used to index as a length n vector of `[row, colum]` indexes. Therefore, the above matrix, will return elements `[2, 4], [4, 3], [7, 1]`:

```{r}
mat[idm]
```
## Lists {#indexlists}

Reminder: A list can contain elements of different class and of different length:

```{r}
(x <- list(one = 1:4,
           two = sample(seq(0, 100, .1), 10),
           three = c("mango", "banana", "tangerine"),
           four = median))
```

You can access a list element with:

- `$` followed by name of the element (therefore only works if elements are named)
- using double brackets `[[` with either name or integer index

To access the third element:

```{r}
x$three

x[["three"]]
class(x[["three"]])
x[[3]]
```

To access an element with a name or integer index stored in a variable, only the bracket notation works - therefore programmatically you would always use double brackets to access different elements:

```{r}
idi <- 3
idc <- "three"
x[[idi]]
x[[idc]]
```

`$` or `[[` return an element.  
In contrast, single bracket `[` indexing of a list returns a pruned list:

```{r}
x[[idi]]
class(x[[idi]])
```
vs.

```{r}
x[idi]
class(x[idi])
```

Extract multiple list elements with single brackets, as expected:

```{r}
x[2:3]
class(x[2:3])
```

Beware (confusing) recursive indexing.  
(This is probably rarely used).  
Unlike in the single brackets example above, colon notation in double brackets accesses elements recursively at the given position.  
The following extracts the 3rd element of the 2nd element of the list:

```{r}
x[[2:3]]
```

You can convert a list to one lone vector containing all the individual components of the original list using `unlist()`. Notice how names are automatically created based on the original structure:

```{r}
(x <- list(alpha = sample(seq(100), 10),
          beta = sample(seq(100), 10),
          gamma = sample(seq(100), 10)))
unlist(x)
```

## Data frames {#indexdfs}

As we've seen, a data frame is a rectangular list. Like a list, its elements are vectors of any type (integer, double, character, factor, and more) but, unlike a list, they have to be of the same length. There are different ways to index and slice a data frame.  
Similar to indexing a list, notice that some return a smaller data frame, while others return vectors.  

```{block, type="Info"}
You can index a data frame using all the ways you can index a list and all the ways you can index a matrix.
```  

Let's create a simple data frame:

```{r}
x <- data.frame(Feat_1 = 21:25,
                Feat_2 = rnorm(5),
                Feat_3 = paste0("rnd_", sample(seq(100), 5)))
x
```

### Extract column(s)

Just like in a list, using the `$` operator returns an element, i.e. a **vector**:

```{r}
x$Feat_2
class(x$Feat_2)
```

Accessing a column by name with square brackets, returns a single-column **data.frame**:

```{r}
x["Feat_2"]
class(x["Feat_2"])
```

Again, similar to a list, if you double the square brackets, you access the element within the data.frame, which is a vector:

```{r}
x[["Feat_2"]]
```

Accessing a column by `[row, column]` either by position or name, return a vector by default:

```{r}
x[, 2]
class(x[, 2])
x[, "Feat_2"]
class(x[, "Feat_2"])
```

The above happens, because by default the argument `drop` is set to `TRUE`. Set it to `FALSE` to return a `data.frame`:

```{r}
class(x[, 2, drop = FALSE])
class(x[, "Feat_2", drop = FALSE])
```

As in lists, with the exception of the `$` notation, all other indexing and slicing operations work with a variable holding either a column name of or an integer location:

```{r}
idi <- 2
idc <- "Feat_2"
x[idi]
x[idc]
x[[idi]]
x[[idc]]
x[, idi]
x[, idc]
x[, idi, drop = F]
x[, idc, drop = F]
```

Extracting multiple columns returns a data frame:

```{r}
x[, 2:3]
class(x[, 2:3])
```

### Extract rows

A row is a small data.frame, since it contains multiple columns:

```{r}
x[1, ]
class(x[1, ])
```

Convert into a list using `c()`:

```{r}
c(x[1, ])
class(c(x[1, ]))
```

Convert into a (named) vector using `unlist()`:

```{r}
unlist(x[1, ])
class(unlist(x[1, ]))
```

## Logical <-> Integer indexing

There are two types of indexes/indices: integer and logical.  

```{block, type="Note"}
- A logical index needs to be of the same dimensions as the object it is indexing (unless you really want to recycle values - see chapter on [vectorization](#vectorization)):  
    you are specifying whether to include or exclude each element
    
- An integer index will be shorter than the object it is indexing:
    you are specifying which subset of elements to include (or with a `-` in front, which elements to exclude)
```

It's easy to convert between the two types of indexes.  

Build an index of all cases in the iris dataset whose sepal length is greater than the mean sepal length:

```{r}
index <- iris$Sepal.Length > mean(iris$Sepal.Length)
head(index)
```

```{r}
head(iris[index, ])
```

### Logical to integer index with `which()`:

`which()` literally gives the position of all `TRUE` elements in a vector, thus converting a logical to an integer index:

```{r}
index_i <- which(iris$Sepal.Length > mean(iris$Sepal.Length))
head(index_i)
table(index)
length(index_i)
```

```{r}
head(iris[index_i, ])
```

### Integer to logical index

```{r}
index_l <- vector(length = NROW(iris))
index_l[index_i] <- TRUE
```

## Complement of an index

To get the complement, you negate the logical index (`!`) and you subtract the integer (`-`):

```{r}
all(iris[!index_l, ] == iris[-index_i, ])
```

## `subset()`

`subset()` allows you to filter cases that meet certain conditions using the `subset` argument, and optionally also select columns using the `select` argument:  

(`head()` returns the first few lines of a data frame. We use it to avoid printing too many lines)

```{r}
head(iris)
iris_sl.gt.med <- subset(iris, Sepal.Length > median(Sepal.Length))
```

Note: You can use the column name Sepal.Length directly, i.e. unquoted and you don't need to use `iris$Sepal.Length`. (This is called Non-Standard Evaluation, NSE)

```{r}
x <- data.frame(one = 1:10,
                two = rnorm(10),
                group = c(rep("alpha", 4),  rep("beta", 6)))
subset(x, subset = two > 0, select = two)
subset(x, two > 0, -one)
subset(x, two > 0, two:one)
subset(x, two > 0, two:group)
```

## `split()`

Split a data frame into multiple data frames by groups defined by a factor:

```{r}
x_by_group <- split(x, x$group)
```

## `with()`

Within a `with()` expression, you can access data.frame columns without quoting or using the `$` operator:

```{r}
with(x, x[group == "alpha", ])
with(x, x[two > 0, ])
```
