# Vectorized Operations {#vectorization}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE) 
```

Most builtin R functions are vectorized and many functions from external packages are as well.

```{block, type="Info"}
A vectorized function operates on all elements of a vector at the same time.
```

Vectorization is very efficient: it can save both human (your) time and machine time.  
In many cases, applying a function on all elements simultaneously may seem obvious or expected behavior, but since not all functions are vectorizes, make sure to check the documentation if unsure.

## Operations between vectors of equal length

```{r}
x <- 1:10
z <- 11:20
x
z
x + z
```

## Operations between a vector and a scalar

```{r}
(x + 10)
(x * 2)
(x / 10)
(x ^ 2)
```

## Operations between vectors of unequal length: value recycling

Operations between a vector and a scalar are a special case of operations between vectors of unequal length. Whenever you perform an operation between two objects of different length, the shorter object's elements are recycled:

```{r}
x + c(2:1)
```

```{block, type = "Note"}
Operations between objects of unequal length can occur by mistake. If the shorter object's length is a multiple of the longer object's length, there will be *no error or warning*, as above. Otherwise, there is a warning, which may be confusing at first, but recycling still happens: ***this is very unlikely to be intentional***:
```

```{r}
x + c(1, 3, 9)
```

## Vectorized matrix operations

Operations between matrices are similarly vectorized.

```{r}
a <- matrix(1:4, 2)
b <- matrix(11:14, 2)
a
b
a + b
a * b
a / b
```

## Vectorized functions

```{r}
log(x)
sqrt(x)
sin(x)
cos(x)
```

## `ifelse()`

`ifelse()` is vectorized:

```{r}
a <- 1:10
(y <- ifelse(a > 5, 11:20, 21:30))
```

but what did this do?

it's equivalent to:

```{r}
idl <- a > 5
yes <- 11:20
no <- 21:30
out <- vector("numeric", 10)
for (i in seq(a)) {
  if (idl[i]) {
    out[i] <- yes[i]
  } else {
    out[i] <- no[i]
  }
}
out
```
i.e.

* Create a logical index using `test`
* for each element `i` in `test`:
    * if the element `i` is TRUE, return `yes[i]`, else `no[i]`
