# Crash Course {#crashcourse}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 5,
                      comment = NA, cache = TRUE) 
```

```{r, comment="", results="asis", echo=FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE)
```

## Introduction & Format

This is a brief introduction to the R programming language for health data science.
It covers basic commands to allow you to read in data, perform common manipulations, plot data and run common tests.

[R](https://en.wikipedia.org/wiki/R_(programming_language)) is a [programming language](https://en.wikipedia.org/wiki/Programming_language) developed specifically for statistical computing and graphics.

It is often mis-characterized as a "statistical package", similar to SPSS, for example, but as a (complete) programming language it has far more extensive functionality.

For a more thorough coverage of the topic, see [Programming for Data Science in R](https://class.lambdamd.org/pdsr/) book.
Links to the book will be provided throughout these notes for those interested in reading up further into particular topics.

R can be used to perform most, if not all, operations in statistics, data science,
machine learning. This is not a crash course in statistics, data science, or 
machine learning, but an introduction to the language itself.

If you have any questions, write them down and make sure to ask them during each section's Q&A.

## Introduction to Programming (!!)

**Everyone can code**

Everyone can learn how to program.
Whether it takes you a minute or a lot longer to learn how to perform a certain
operation using code, you can master it. Do not worry about comparing yourself 
to others.

As with more or less everything, motivation is key.

You are here because, presumably, you have important questions to answet using data.

Knowing even a little code can give you the power to work with your own data without
depending fully on someone else. At the same time, it makes collaborating with
other scientistics, clinicians, statisticians, etc. much more effective & efficient.

Learning to code can be exciting and frustrating. It's similar to learning to
play an instrument - like the guitar. At first, it may seem unnatural and
annoying, but you can get better at it rather quickly and it's very rewarding 
and satisfying.

You don't learn to code by reading books or slides, you learn to code when you
**practice, practice, practice**. Reading other people's code can be very helpful.
It is important to **be able to read & write code**.

**Coding needs logic & creativity**

Programming is based on logic and learning to code helps structure your thoughts,
your experiments, your reasoning.

Programming languages are human constructs. They developed to answer important needs.
They develop with time, as needs evolve. Many design choices are explained historically,
a few may be more arbitrary.

**Everything is a series of simple, manageable steps**

Remember this: the shortest and simplest piece of code up to the longest and most complex
is made of a sequence of relatively simple steps.

Always be clear about what you want to achieve first, then break it down step-by-step
in code. Each step is relatively easy to figure out and **check** that it is working
as desired. A common mistake is to write multiple steps in one go, and then have 
a hard time figuring out where an error occurs or why.

To get from A to B using code there are virtually always multiple different paths.
That can be confusing or perhaps frustrating, but it is also exciting.
Programming to the uninitiated may seem a a rigid exercise but it is highly creative.
Remember that there objective and subjective differences to consider when designing
a code to take you from A to B. Suppose you have two approaches that have the same input 
and produce the same output. An objective difference would be how fast each completes the
task and how many lines of code or number of function calls it requires. A subjective
difference would be the programming style / syntax used / whether the code is 
"elegant" - a pretty broad topic.

**Always check yourself**

Always remember: the most important thing is to ensure you have 
correct results at each step. Don't place writing smart or elegant code above
writing correct code. 

**Errors happen and they are not all the same**

Errors in code happen all the time, it is part of the process. But, not all errors
are the same, far from it. One crucial difference is coding errors that:

- **stop execution** of code and **produce an error message**. This is the best case scenario
because it can't go unnoticed.
- **do not stop execution** of code but **produce a warning**. These warnings are
too often ignored. They may be serios or trivial, but must be investigated.
- **do not stop execution** and **produce no warnings**. This is the worst kind of error
since it is silent. These are very common and the only way to recognize them are
to check the output. 

**Details matter (a lot)**

A lot of beginner and non-beginner mistakes occur because
a variable or function name is misspelled.

ALWAYS, ALWAYS, ALWAYS READ ERROR AND WARNING MESSAGES.
The all caps is because this is a) essential and b) far too often ignored.

**Document everything**

Use comments in your code to explaing what you are trying to achieve and why.
You will often need to revisit your code after some time has passed. Life will 
be very hard if it's not clear what is happening and why. 

Programming is largely a team sport. A lot of code is written collaboratively
or is used by people other than the author. Again, comprehensive documentation 
is super important.

**Help is at your fingertips**

Whether you are just starting out or you are a seasoned programmer, you have
many sources of information to help you troubleshoot or learn new skills.

- Use the built-in documentation! Builtin help files, written by the code author,
are almost always the best place to start. Their quality will vary, but they
are often sufficient to learn how to use a function properly.
- Programming is an online activity. Most errors or difficulties you encounter
have been encountered many times before by others. Millions of Q&A sites, blogs,
forums are a web search away. Copy-pasting an error message into a search engine 
will often result in multiple hits.

## The R language

- The **S** statistical programming language was developed in 1976 at Bell Labs by John Chambers and others "to turn ideas into software, quickly and faithfully”.

- **R** is an [open source](https://en.wikipedia.org/wiki/Open_source) implementation of S developed by Ross Ihaka and Robert Gentleman at the University of Auckland, New Zealand; initial version released in 1995.

- Supported by the [R Foundation for Statistical Computing](https://www.r-project.org/foundation/), developed by the [R Core Team and the contributions of many others](https://www.r-project.org/contributors.html).

- Official part of the [Free Software Foundation’s](https://www.fsf.org) [GNU project](https://www.gnu.org/) available under [GNU GPL v2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.html).

- Latest version [4.1.3 released 2022-03-10](https://cloud.r-project.org/) (as of writing)

## Free Open Source Software

[Free Open Source Software](https://en.wikipedia.org/wiki/Free_and_open-source_software)

[What is Free Software?](https://www.fsf.org/about/what-is-free-software)

> "Free software is software that gives you the user the freedom to share, 
>  study and modify it. We call this free software because the user is free."
> 
> `r tufte::quote_footer('--- Free Software Foundation')`

[What is Open Source Software](https://opensource.com/resources/what-open-source)

> "Open source software is software with source code that anyone can 
> inspect, modify, and enhance."
>
> `r tufte::quote_footer('--- opensource.com')`

Free and Open Source Software is importance in biomedical research and clinical applications
for a number of reasons, including promoting **inclusion**, **transparency**, and **trustworthiness.**

## RStudio Integrated Development Environment

[RStudio](https://www.rstudio.com/) offers a popular, feature-full [Integrated Development Environment (IDE)](https://en.wikipedia.org/wiki/Integrated_development_environment) for R.

More advanced users can use [Visual Studio Code](https://code.visualstudio.com/) with the [R Extension](https://github.com/REditorSupport/vscode-R) for a similar R IDE experience together with all the extra functionality an convenience of VS Code.

## The R core language & package ecosystem

R boasts extensive quantitative and statistical functionality in the base system.

This functionality is extended through a vast ecosystem of external packages.

- **CRAN**: The Comprehensive R Archive Network: https://cran.r-project.org/
- **Bioconductor**: Bioinformatics-related packages and more: https://www.bioconductor.org/
- **GitHub**: The largest source code host (>200M repositories): https://github.com/

## Reading in Data

We shall use a heart failure dataset as an example. It is freely available at the UCI repository:
"https://archive.ics.uci.edu/ml/machine-learning-databases/00519/heart_failure_clinical_records_dataset.csv"

### CSV

```{r}
dat <- read.csv("~/icloud/Data/UCI/heart_failure_clinical_records_dataset.csv")
```

The `head()` function prints the first few lines of an object:

```{r}
head(dat)
```

The `read.csv()` function read the contents of the CSV file into an R object known
as a `data.frame`. This is essentially a table like a spreadsheet, where each row
represents a case (e.g. a subject, patient, etc.) and each columnn represents a
variable (e.g. Patient ID, Age, Sex, Dx, etc.)

### XLSX

```{r}
dat_too <- openxlsx::read.xlsx("~/icloud/Data/UCI/heart_failure_clinical_records_dataset.xlsx")
```

```{r}
head(dat_too)
```

## Inspect & summarize data

Get data dimensions:

```{r}
dim(dat)
```

Look at the data structure, including data types:

```{r}
str(dat)
```

Get summary of dataset:

```{r}
summary(dat)
```

## Data types

R includes a number of builtin data types. Some of the most common are:

- numeric (e.g. 1.2, 5.9, 11.4)
- character (e.g. "SF", "SD")
- logical (e.g. "TRUE", "FALSE")

## Data Structures

R includes multiple different data structures. Think of a data structure as a
container that holds one or more vectors of data.

The data.frame is one of the most common data structures for statistics, because
it can hold vectors of different kinds, e.g. numeric, categorical, and character.

```{block, type="rmdnote"}
[Read more about data structures in PDSR](https://class.lambdamd.org/pdsr/datastructures.html)
```

### Factors

Factors in R are used to store categorical variables and therefore have many 
important uses in statistics / data science / machine learning.

Let's convert binary categorical variables in our dataset to factors:

```{r}
dat$anaemia <- factor(dat$anaemia)
dat$diabetes <- factor(dat$diabetes)
dat$high_blood_pressure <- factor(dat$high_blood_pressure)
dat$sex <- factor(dat$sex)
dat$smoking <- factor(dat$smoking)
dat$DEATH_EVENT <- factor(dat$DEATH_EVENT)
```

```{block, type="rmdnote"}
[Read more about factors in PDSR](https://class.lambdamd.org/pdsr/factors.html)
```

## Summarize data

A lot of statistical functionality is built in to the language. You can easily get
summary statistics of variables using functions like `mean()`, `median()`, `range()`,
`max()`, `min()`.

To access a specific column of a data.frame you use the `$` notation.
We use `head()` to avoid printing the entire variable.

```{r}
head(dat$age)
```

### Continuous variables

```{r}
mean(dat$age)
median(dat$age)
min(dat$age)
max(dat$age)
summary(dat$age)
```

### Categorical variables

The `table()` function gives you the counts for each level/unique value of a variable:

```{r}
table(dat$sex)
table(dat$smoking)
```

## Plots

R has powerful and extensive support for graphics built in to the core language.

Here, we look at how to produce some common and important plot types:

### Boxplot

```{r}
boxplot(dat$serum_sodium ~ dat$smoking, col = "royalblue")
```

### Histogram

```{r}
hist(dat$age, col = "royalblue")
```

### Scatter plot

```{r}
plot(dat$age, dat$serum_sodium, col = "royalblue")
```

## Hypothesis Testing

```{r}
chisq.test(dat$smoking, dat$DEATH_EVENT)
```

```{r}
smoking_sex <- chisq.test(dat$smoking, dat$sex)
smoking_sex
```
```{r}
smoking_sex$expected
```

```{r}
smoking_sex$observed
```

## Saving data

### CSV

You can write R objects to CSV file using `write.csv()'

```{r eval = FALSE}
write.csv(dat, "~/Data/dat.csv")
```

### RDS

You can also directly save any R object as an "RDS" file

```{r eval = FALSE}
saveRDS(dat, "~/Data/dat.rds")
```
