# Aggregate {#aggregate}

`aggregate()` is a powerful way to apply functions on splits of your data. It can replicate 
functionality of the `*apply()` family, but can be more flexible. This may come with a performance
penalty, only noticeable with big data, in which case it is recommended to use 
[data.table](#datatable) for fast group-by data summarization.

`aggregate()` can work either with a formula notation or directly on data.frames and vectors. We 
show how to perform each operation below with either approach. The formula interface might be easier 
to work with interactivly on the console. While you *can* code with the formula interface, the 
regular approach is a lot more straightforward to do so.

For this example, we get the penguin data:

```{r}
library(palmerpenguins)
str(penguins)
```

See example below for 1 or multiple variables by 1 or more groups using either the formula 
interface, or working directly on objects with `$` indexing or using `with()`:

## Single variable by single group

Note that the formula method defaults to na.action = na.omit

Using the formula interface:

```{r}
aggregate(bill_length_mm ~ species,
          penguins, mean)
```

Directly working with vectors:

```{r}
aggregate(penguins$bill_length_mm,
          by = list(penguins$species),
          mean, na.rm = T)
```

Using `with()`:

```{r}
with(penguins,
     aggregate(bill_length_mm,
               by = list(species),
               mean, na.rm = TRUE))
```

## Multiple variables by single group

```{r}
aggregate(cbind(bill_length_mm, flipper_length_mm) ~ species,
          penguins, mean)
```

```{r}
aggregate(penguins[, c("bill_length_mm", "flipper_length_mm")],
          by = list(penguins$species),
          mean, na.rm = TRUE)
```

```{r}
with(penguins,
     aggregate(cbind(bill_length_mm, flipper_length_mm),
               by = list(species),
               mean, na.rm = TRUE))
```

## Single variable by multiple groups

```{r}
aggregate(bill_length_mm ~ species + island, penguins, mean)
```

```{r}
aggregate(penguins$bill_length_mm,
          by = list(penguins$species, penguins$island),
          mean, na.rm = TRUE)
```

```{r}
with(penguins,
     aggregate(bill_length_mm,
               by = list(species, island),
               mean, na.rm = TRUE))
```

## Multiple variables by multiple groups

```{r}
aggregate(cbind(bill_length_mm, flipper_length_mm) ~ species + island,
          penguins, mean)
```

```{r}
aggregate(penguins[, c("bill_length_mm", "flipper_length_mm")],
          by = list(penguins$species, penguins$island),
          mean, na.rm = TRUE)
```

```{r}
with(penguins,
     aggregate(cbind(bill_length_mm, flipper_length_mm),
               by = list(species, island),
               mean, na.rm = TRUE))
```

## See also

[`tapply()`]({#tapply}) for an alternative methods of applying function on subsets of a single 
variable (probably faster)
