# Summarizing Data {#summarize}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5,
                      comment = NA, cache = TRUE)
```

```{r, comment="", results="asis", echo=FALSE}
old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
options(crayon.enabled = TRUE)
```

Let's read in a dataset from OpenML:

```{r}
heart <- read.csv("https://www.openml.org/data/get_csv/51/dataset_51_heart-h.arff",
                  na.strings = "?")
```


## Get summary of an R object with `summary()`

R includes `summary()` [methods](#s3methods) for a number of different objects. 

```{r}
summary(heart)
```

## Fast builtin column and row operations

We saw in [Loop Functions](#loopfns) how we can apply functions on rows, columns, or other subsets of our data. R has optimized builtin functions for some very common operations, with self-explanatory names:

```{r}
a <- matrix(1:20, 5)
a
```

```{r}
colSums(a)

# same as
apply(a, 2, sum)
```

```{r}
rowSums(a)

# same as
apply(a, 1, sum)
```

```{r}
colMeans(a)

# same as
apply(a, 2, mean)
```

```{r}
rowMeans(a)

# same as
apply(a, 1, mean)
```

## Optimized matrix operations with **matrixStats**

Whilte the builtin operations above are already optimized and faster than the equivalent calls, the **matrixStats** package [@matrixStats2019] offers a number of futher optimized matrix operations, including drop-in replacements of the above. These should be prefered when dealing with bigger data:

```{r}
library(matrixStats)
colSums2(a)
rowSums2(a)
colMeans2(a)
rowMeans2(a)
```

Note: **matrixStats** provides replacement functions named almost identically to their base counterpart - so they are easy to find - but are different - so they don't mask the base functions (this is important and good software design).

## Grouped summary statistics with `aggregate()`

`aggregate()` is a powerful way to apply functions on splits of your data. It can replicate functionality of the `*apply()` family, but can be more flexible/powerful and supports a formula input.

```{r}
aggregate(iris[, -5], by = list(iris$Species), mean)
```

Alternatively, you can use the very "R", very convenient, **formula notation**:

```{r}
aggregate(. ~ Species, iris, mean)
```

Let's make up a second grouping:

```{r}
irisd <- iris
irisd$Group2 <- rep(1:2, 3, each = 25)
irisd$Group2
```

This "Group2" divides each iris Species in the first and last 25 cases.  
Let's aggregate by both Species and Group2:

```{r}
aggregate(irisd[, -c(5, 6)],
          by = list(Species = irisd$Species, Group2 = irisd$Group2),
          mean)
```

And using the more compact formula notation:
```{r}
aggregate(. ~ Species + Group2, irisd, mean)
```

Note: Using aggregate with the `by = list()` argument is easier to code with. The formula notation might be easier to work with in real time on the console. You *can* code with the formula notation, but if there is an alternative it's unlikely to be worth the extra steps.
