# Intro to data modeling {#modeling}

```{r echo = FALSE}
knitr::opts_chunk$set(fig.width = 4.5, fig.height = 4.5,
                      comment = NA, cache = TRUE) 
```
-> This section is a work in progress

## Generalized Linear Model (GLM)

The Generalized Linear Model is one of the most common and important models in statistics.  
Let's look at an example using the GLM for regression.  
We will use the `mtcars` builtin dataset to predict horsepower (`hp`) of 32 cars from 10 other features:

```{r}
str(mtcars)
```
Here, we use the `glm()` function which accepts a formula that defines the model.  
The formula used below means "regress hp on all other variables". The `family` argument defines we are performing regression and the `data` argument points to the data frame where the covariates used in the formula are found.

```{r}
mod <- glm(hp ~ ., family = "gaussian", data = mtcars)
mod
```

For amgaussian output, we can also use the `lm()` function (there are minor differences in the output created, but the model is the same):

```{r}
mod <- lm(hp ~ ., data = mtcars)
mod
```

Get summary of the model using `summary()`:

```{r}
summary(mod)
```

Note how R prints stars next to covariates whose p-values falls within certain limits, described right below the table of estimates.  
Above, for example, the p-value for `disp` falls between 0.001 and 0.01 and therefore gets highlighted with 2 stars.  

To extract the p-values of the intercept and each coefficient, we use `coef()` on `summary()`. The final (4th) column lists the p-values:

```{r}
coef(summary(mod))
```

## Mass-univariate analysis

There are many cases where we have a large number of predictors and, along with any other number of tests or models, we may want to regress our outcome of interest on each covariate, one at a time.  

Let's create some synthetic data with 1000 cases and 100 covariates  
The outcome is generated using just 4 of those 100 covariates and has added noise.

```{r}
set.seed(2020)
n_col <- 100
n_row <- 1000
x <- as.data.frame(lapply(seq(n_col), function(i) rnorm(n_row)),
                   col.names = paste0("Feature_", seq(n_col)))
dim(x)
y <- .7 + x[, 10] + .3 * x[, 20] + 1.3 * x[, 30] + x[, 50] + rnorm(500)
```

Let's fit a linear model regressing y on each column of x using `lm`:

```{r}
mod.xy.massuni <- lapply(seq(x), function(i) lm(y ~ x[, i]))
length(mod.xy.massuni)
names(mod.xy.massuni) <- paste0("mod", seq(x))
```

To extract p-values for each model, we must find where exactly to look.  
Let's look into the first model:
```{r}
(ms1 <- summary(mod.xy.massuni$mod1))
ms1$coefficients
```

The p-values for each feature is stored in row 1, column 4 fo the coefficients matrix. Let's extract all of them:
```{r}
mod.xy.massuni.pvals <- sapply(mod.xy.massuni, function(i) summary(i)$coefficients[2, 4])
```
Let's see which variable are significant at the 0.05:
```{r}
which(mod.xy.massuni.pvals < .05)
```
...and which are significant at the 0.01 level:
```{r}
which(mod.xy.massuni.pvals < .01)
```

## Multiple comparison correction

We've performed a large number of tests and before reporting the results, we need to control for [multiple comparisons](https://en.wikipedia.org/wiki/Multiple_comparisons_problem).  
To do that, we use R's `p.adjust()` function. It adjusts a vector of p-values to account for multiple comparisons using one of multiple methods. The default, and recommended, is the [Holm method](https://en.wikipedia.org/wiki/Holm%E2%80%93Bonferroni_method). It ensures that `FWER < α`, i.e. controls the [family-wise error rate](https://en.wikipedia.org/wiki/Family-wise_error_rate), a.k.a. the probability of making one or more false discoveries (Type I errors) 

```{r}
mod.xy.massuni.pvals.holm_adjusted <- p.adjust(mod.xy.massuni.pvals)
```

Now, let's see which features' p-values survive the magical .05 threshold:

```{r}
which(mod.xy.massuni.pvals.holm_adjusted < .05)
```

These are indeed the correct features (not surprisingly, still reassuringly).
