---
execute:
  warning: false
knitr:
  opts_knit:
    global.par: true
  opts_chunk:
    comment: ''
    fig.width: 5.5
    fig.height: 5.5
    out.width: 80%
---

# Introduction to **rtemis.draw** {#sec-rtemisdraw}

**`rtemis.draw`** is a plotting package from the [rtemis](https://rtemis.org) project. Like **`plotly`** (@sec-plotly), it produces interactive
[htmlwidgets](https://www.htmlwidgets.org), but it provides data-first `draw_*()` functions
that take plain vectors, lists, or data frames and choose sensible defaults, so most plots
take a single call. Charts are rendered with
[Apache ECharts](https://echarts.apache.org), and automatically follow the light or dark
theme of the page they are displayed in.

::: callout-note
`rtemis.draw` is available from [R-universe](https://rtemis-org.r-universe.dev) and is currently
under review for CRAN.
:::

Full documentation is available at [docs.rtemis.org](https://docs.rtemis.org/r/draw/).

## Setup

### Installation

Until the package is released on CRAN, install it from R-universe:

```{r}
#| eval: false
install.packages(
    "rtemis.draw",
    repos = c("https://rtemis-org.r-universe.dev", "https://cloud.r-project.org")
)
```

### Libraries

Load libraries we will be using in this chapter:

```{r}
library(rtemis.draw)
library(data.table)
options(datatable.print.class = TRUE)
```

### Synthetic Data

We use the same synthetic dataset as in the previous two chapters:

```{r}
set.seed(2022)
dt <- data.table(
    PID = sample(8001:9000, size = 100),
    Age = rnorm(100, mean = 33, sd = 8),
    Weight = rnorm(100, mean = 70, sd = 9),
    SysBP = rnorm(100, mean = 110, sd = 6),
    DiaBP = rnorm(100, mean = 80, sd = 6),
    Sex = factor(sample(c("Female", "Male"), size = 100, replace = TRUE))
)
dt[, SysBP := SysBP + 0.5 * Age]
dt[Sex == "Male", Weight := Weight + rnorm(.N, mean = 16, sd = 1.5)]
dt[Sex == "Male", Age := Age + rnorm(.N, mean = 6, sd = 1.8)]
```

Define a color palette `palette`. Unlike in the previous chapters, we do not need a separate
semi-transparent version of the palette: functions that fill areas, like `draw_histogram()` and
`draw_density()`, have a `fill_alpha` argument, which is automatically set depending on whether there are multiple groups or not, but can be overridden if desired.

```{r}
palette <- c("#43A4AC", "#FA9860")
```

::: callout-note
All `draw_*()` functions share a common set of arguments: `title`, `xlab`, `ylab`, `theme`, `width`,
`height`, `palette`, and `filename`. If `palette` is not specified, the rtemis palette is used.
:::

## Box plot

`draw_boxplot()` accepts a numeric vector, a list of numeric vectors, or a data frame
of numeric columns, and draws one box per element or column. Unlike **`ggplot2`**, there is no need
to convert to long format first. Unlike **`plotly`**, there is no need to add each box one at a time.

```{r}
draw_boxplot(dt[, .(SysBP, DiaBP)])
```

Use `palette` to specify the colors and `labels` to specify the box labels:

```{r}
draw_boxplot(
    dt[, .(SysBP, DiaBP)],
    palette = palette,
    labels = c("Systolic BP", "Diastolic BP")
)
```

### Grouped boxplot

Use `group` to split a vector by a factor:

```{r}
draw_boxplot(dt[, Age], group = dt[, Sex], palette = palette[2:1])
```

## Histogram

```{r}
draw_histogram(dt[, Age], palette = palette[1])
```

We can specify the number of bins with `breaks`, and axis labels with `xlab` and `ylab`:

```{r}
draw_histogram(
  dt[, Age],
  breaks = 24,
  palette = palette[1],
  xlab = "Age (years)"
)
```

### Grouped Histogram

```{r}
draw_histogram(
  dt[, Age],
  group = dt[, Sex],
  palette = palette[2:1]
)
```

`bar_mode` controls the placement of the groups' bars within each bin: `"group"` (the default)
places them side by side, whereas `"stack"` stacks them vertically.

```{r}
draw_histogram(
  dt[, Age],
  group = dt[, Sex],
  breaks = 24,
  bar_mode = "stack",
  palette = palette[2:1]
)
```

## Density plot

`draw_density()` estimates the density internally with `stats::density()`, so there is no need to
compute and pass the coordinates ourselves as with **`plotly`**.

```{r}
draw_density(dt[, Age], palette = palette[1])
```

### Grouped density plot

```{r}
draw_density(dt[, Age], group = dt[, Sex], palette = palette[2:1])
```

## Barplot

```{r}
schools <- data.frame(UCSF = 4, Stanford = 7, Penn = 12)
```

`draw_bar()` takes the category labels as `x` and the bar heights as `y`. No data frame
or explicit factor is needed; the bars appear in the order provided:

```{r}
draw_bar(
    names(schools),
    as.numeric(schools[1, ]),
    palette = palette[1],
    ylab = "N schools"
)
```

## Scatterplot

```{r}
draw_scatter(dt[, Age], dt[, SysBP], xlab = "Age", ylab = "SysBP", palette = palette[1])
```

Pass `fit = "gam"` (or `"glm"`) to overlay a fitted line and 95% confidence band:

```{r}
draw_scatter(
    dt[, Age], dt[, SysBP],
    fit = "gam",
    xlab = "Age", ylab = "SysBP",
    palette = palette[1]
)
```

### Grouped Scatterplot

```{r}
draw_scatter(
    dt[, Age], dt[, SysBP],
    group = dt[, Sex],
    xlab = "Age", ylab = "SysBP",
    palette = palette[2:1]
)
```

With a grouping variable, `fit` draws a separate fitted line for each group:

```{r}
draw_scatter(
    dt[, Age], dt[, SysBP],
    group = dt[, Sex],
    fit = "gam",
    xlab = "Age", ylab = "SysBP",
    palette = palette[2:1]
)
```

## Themes

Charts follow the page theme (light or dark) by default. To force a theme, pass one to the
`theme` argument:

```{r}
draw_scatter(dt[, Age], dt[, SysBP], theme = theme_dark())
```

## Save plot to file

We'll use the grouped boxplot example from above to show how to save a plot to file. Pass an `.svg`
`filename` to any `draw_*()` function, or call `save_drawing()` on an existing widget.
SVG export requires [Node.js](https://nodejs.org).

```{r}
#| eval: false
p <- draw_boxplot(dt[, Age], group = dt[, Sex], palette = palette[2:1])
save_drawing(p, "Age_by_Sex_rtemis.draw.svg", width = 550, height = 550)
```

or directly:

```{r}
#| eval: false
draw_boxplot(
    dt[, Age], group = dt[, Sex], palette = palette[2:1],
    filename = "Age_by_Sex_rtemis.draw.svg"
)
```

## See also

- Base R Graphics ([@sec-basegraphics])
- Introduction to **ggplot2** ([@sec-ggplot2])
- Introduction to **plotly** ([@sec-plotly])

## Resources

- [rtemis.draw Documentation](https://docs.rtemis.org/r/draw/)
