---
knitr:
  opts_chunk: 
    comment: ''
---
# Basic operations {#basicops}

First, before even learning about data types and structures, it may be worth 
looking at some of the basic mathematical and statistical operations in R.

## Mathematical operations

We use the assignment operator `<-` to assign values to variables `x` and `y`:

```{r}
x <- 10
y <- 3
```

Standard arithmetic operations are as expected:

```{r}
x + y
x - y
x * y
x / y
```
Exponentiation uses `^`:
(The caret (`^`) is likely the most common but not the only symbol used for 
exponentiation across programming languages)

```{r}
x^3
```

Square root is `sqrt()`:

```{r}
sqrt(81)
```

Natural logs with `log()`:

```{r}
log(12)
```

Base 10 log with `log10()`:

```{r}
log10(1000)
```

Exponential function with `exp()`:

```{r}
exp(1)
```

Integer division i.e. *Divide and forget the remainder*

```{r}
x %/% y
```

i.e. how many times the denominator fits in the numerator, without taking 
fractions of the denominator. It can be applied on decimals the same way:

```{r}
9.5 %/% 3.1
```

Modulo operation i.e. *Divide and return just the remainder*

```{r}
x %% y
```

```{r}
x <- (-10:10)[-11]
y <- sample((-10:10)[-11], size = 20)
x - (x %/% y) * y == x %% y
```

Trigonometric functions

```{r}
cos(x)
sin(x)
tan(x)
```

See `?cos` for more specialized trigonometric functions and details.

## Logical operations

Logical AND: `&`

```{r}
TRUE & TRUE
```

```{r}
TRUE & FALSE
```

Logical OR: `|`

```{r}
TRUE | FALSE
```

Logical negation: `!`

```{r}
x <- TRUE
!x
```

Exclusive OR: `xor()`  
XOR evaluates to TRUE when two logicals are different,  
i.e. one or the other is TRUE but not both.

```{r}
a <- c(TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)
b <- c(FALSE, FALSE, TRUE, FALSE, TRUE, TRUE)
a & b
a | b
xor(a, b)
```

Test all elements of an object are TRUE with `all()`:

```{r}
all(a)
```

Test if any element is TRUE with `any()`:

```{r}
any(a)
```

## Common statistical operations

First, let's use the `rnorm()` function to draw 200 numbers at random from a 
normal distribution:

```{r}
x <- rnorm(200)
```

### Descriptive statistics

mean, median, standard deviation, minimum, maximum, and range:

```{r}
mean(x)
median(x)
sd(x)
min(x)
max(x)
range(x)
```

```{r}
summary(x)
```

### Sampling

R's `sample()` allows you to samples elements of an 
[R object](#sec-datastructures) with or without replacement:

By default, the `replace` argument is set to `FALSE`, i.e. sampling is 
performed without replacement and you can request a sample size up to the
length of the object:

```{r}
x <- 21:30
sample(x, size = 5)
```

`size` is the second argument and its name can therefore be omitted if it is the
second value you pass to `sample()`:

```{r}
sample(x, 10)
```

Setting `replace = TRUE` performs sampling with replacement and you can set 
size to any positive integer, including values larger than the length of the
input:

```{r}
sample(x, size = 100, replace = TRUE)
```
