---
execute:
  cache: true
knitr:
  opts_chunk: 
    comment: ''
---
# Basic operations {#basicops}

First, before even learning about data types and structures, it may be worth 
looking at some of the basic mathematical and statistical operations in R.

## Arithmetic

```{r}
x <- 10
y <- 3
```

Standard arithmetic operations are as expected:

```{r}
x + y
x - y
x * y
x / y
```
Exponentiation uses `^`:
(The caret (`^`) is likely the most common but not the only symbol used for exponentiation across programming languages)

```{r}
x^3
```

Square root is `sqrt()`:

```{r}
sqrt(81)
```

Natural logs with `log()`:

```{r}
log(12)
```

Base 10 log with `log10()`:

```{r}
log10(1000)
```

Integer division i.e. *Divide and forget the remainder*

```{r}
x %/% y
```

i.e. how many times the denominator fits in the numerator, without taking fractions of the 
denominator. It can be applied on decimals the same way:

```{r}
9.5 %/% 3.1
```

Modulo operation i.e. *Divide and return just the remainder*

```{r}
x %% y
```

```{r}
x <- (-10:10)[-11]
y <- sample((-10:10)[-11], 20)
x - (x %/% y) * y == x %% y
```

## Logical operations

Logical AND: `&`

```{r}
T & T
```

```{r}
T & F
```

Logical OR: `|`

```{r}
T | F
```

Logical negation: `!`

```{r}
x <- TRUE
!x
```

Exclusive OR: `xor()`  
XOR evaluates to TRUE when two logicals are different,  
i.e. one or the other is TRUE but not both.

```{r}
a <- c(T, T, T, F, F, F)
b <- c(F, F, T, F, T, T)
a & b
a | b
xor(a, b)
```

Test all elements of an object are TRUE with `all()`:

```{r}
all(a)
```

Test if any element is TRUE with `any()`:

```{r}
any(a)
```

## Common descriptive stats

First, let's use the `rnorm()` function to draw 200 numbers at random from a 
normal distribution:

```{r}
x <- rnorm(200)
```

Basic descriptive stat operations -  
mean, median, standard deviation, minimum, maximum, and range:

```{r}
mean(x)
median(x)
sd(x)
min(x)
max(x)
range(x)
```

```{r}
summary(x)
```
