---
knitr:
  opts_chunk:
    comment: ""
    cache: true
lightbox: true
---

# Advanced Usage {#sec-dtadvanced}

This is a growing list of slightly more advanced examples using **data.table**.

```{r}
library(data.table)
options(datatable.print.class = TRUE)
```

## Count by group using `.N`
```{r}
library(palmerpenguins)
peng <- as.data.table(penguins)
```

Example: Count penguins by island:

```{r}
peng[, .N, by = island]
```

Example: Count by island and species:

```{r}
peng[, .N, by = list(island, species)]
```

## Index within groups using `.I`

For example, get index for min or max within each group, where index
refers to the entire data.table.

Look at the documentation for `.I`

```{r}
ir <- as.data.table(iris)
ir[, .I[which.max(Petal.Width)], by = Species]
```

Use that to get the case with max Petal.Width from each Species

```{r}
ir[, ir[, .I[which.max(Petal.Width)], by = Species][[2]]]
```

## See also

- [data.table docs](https://rdatatable.gitlab.io/data.table/index.html)
