# data.table Data I/O {#sec-dtdataio}

## Read delimited data with `fread()`

**data.table**'s "fast read" function `fread()` offers a replacement for the 
base `read.csv()` command. By default, it creates a data.table object, which
is likely preferred in most scenarios, but it can create a data.frame by 
setting `data.table = FALSE`.

It is very useful because it allows:

- very fast, parallelized reading of large delimited files, e.g. with millions 
of rows
- can directly read zipped files
- smart automatic discovery of delimiters when `sep = "auto"` (default)
- multiple other conveniences

As an example, you can read a gzipped file directly from a URL:

```{r}
library(data.table)
```

```{r}
dat <- fread("https://archive.ics.uci.edu/ml/machine-learning-databases/00519/heart_failure_clinical_records_dataset.csv")
```

```{r}
dat
```

### See also

[Convenience features of fread](https://github.com/Rdatatable/data.table/wiki/Convenience-features-of-fread)

## Write delimited data with `fwrite()`

`fwrite()` similarly provides a faster, parallelized, and more flexible replacement for 
`write.csv()`:

```{r}
#| eval: false
fwrite(dat, "/path/to/file.csv")
```

## Save a `data.table` to file

Same as any R object, you can save a `data.table` to disk using `saveRDS()`.
Suppose you have read data in with `fread()` or coerced a dataset using 
`as.data.table()`, done some cleaning up, type conversions, data 
transformations, etc, this is the preferred way to save your work, so you can 
reload at any time.

```{r}
#| eval: false
saveRDS(dat, "/path/to/data.rds")
```
