---
knitr:
  opts_chunk:
    comment: ""
---
# Filesystem operations

R includes multiple functions to interact with the filesystem. The most
common ones are:

- Get current working directory: `getwd()`
- Set the working directory: `setwd()`
- List files or directories: `list.files()`, `list.dirs()`
- Create file or directory: `file.create()`, `dir.create()`
- Check if file or directory exists: `file.exists()`, `dir.exists()`
- Remove file or directory: `file.remove()`, `dir.remove()`
- Copy file or directory: both using `file.copy()`, use `recursive = TRUE` to copy directory and its contents

## Construct path to file

`file.path()` is a useful function that constructs a platform-specific path to a file or directory. It is useful because it automatically adds the correct path separator (`/` or `\`) depending on the operating system.

```{r}
file.path("Preprocessed", "dat_p.rds")
```

## Expand path

`path.expand()` expands


## Normalize path

`normalizePath()`

```{r}
normalizePath("Preprocessed/dat_p.rds", mustWork = FALSE)
```