---
engine: knitr
knitr:
  opts_chunk:
    comment: ""
lightbox: true
---
# Introduction {#sec-intro}

The [S Programming Language](https://en.wikipedia.org/wiki/S_(programming_language)) is a statistical
programming language developed at [Bell Labs](https://en.wikipedia.org/wiki/Bell_Labs) 
"to turn ideas into software, quickly and faithfully" [@chambers1998programming].

[R](https://en.wikipedia.org/wiki/R_(programming_language)) is a modern implementation of the
S language and part of the [GNU Project](https://en.wikipedia.org/wiki/GNU_Project).

R is an [interpreted](https://en.wikipedia.org/wiki/Interpreter_(computing)) language, allowing
interactive work with data. It is written in C, Fortran, and R itself.

Some of R's strengths:

- The **base** language (i.e. what is included in R when you first install it) 
comes loaded with functionality for:

  - data cleaning and manipulation
  - statistical testing and modeling
  - powerful graphics

- The vast ecosystem of **third party packages** brings extensive 
functionality for statistics, epidemiology, machine learning, visualization, 
image processing and much more. This includes specialized packages for many 
biomedical applications.
<br>

See also: [What is R?](https://www.r-project.org/about.html) on the 
[R Project website](https://www.r-project.org/).
<br>

## Keeping up to date

This book was compiled using `r R.Version()$version.string`.  
Make sure you have the latest version by visiting the 
[R project website](https://www.r-project.org).
<br>

New versions of R are released regularly and it is a good idea to keep your
version of R and installed packages updated.

You can get the version of R currently in use by printing the `version` object:

```{r}
version
```

It's also a good idea to keep a log of the version of R and installed packages 
when beginning a new project. An easy way to do this is to save the output of 
`sessionInfo()`:

```{r}
sessionInfo()
```
