---
execute: 
  warning: false
knitr:
  opts_knit:
    global.par: true
  opts_chunk: 
    comment: ''
    fig.width: 5.5
    fig.height: 5.5
    out.width: 80%
---

# Introduction to **plotly**

[Plotly](https://plot.ly) [@sievert2017plotly] is a comprehensive plotting library that
produces interactive plots.

## Setup

### Libraries

Load libraries we will be using in this chapter:

```{r}
library(plotly)
library(data.table)
options(datatable.print.class = TRUE)
```

### Synthetic Data

```{r}
library(data.table)
set.seed(2022)
dt <- data.table(
    PID = sample(8001:9000, size = 100),
    Age = rnorm(100, mean = 33, sd = 8),
    Weight = rnorm(100, mean = 70, sd = 9),
    SysBP = rnorm(100, mean = 110, sd = 6),
    DiaBP = rnorm(100, mean = 80, sd = 6),
    Sex = factor(sample(c("Female", "Male"), size = 100, replace = TRUE))
)
dt[, SysBP := SysBP + .5 * Age]
dt[Sex == "Male", Weight := Weight + rnorm(.N, mean = 16, sd = 1.5)]
dt[Sex == "Male", Age := Age + rnorm(.N, mean = 6, sd = 1.8)]
```

Define a color palette `palette` and a version of the same palette at 2/3 transparency,
`palette_a`:

```{r}
palette <- c("#43A4AC", "#FA9860")
palette_a <- adjustcolor(palette, alpha.f = 2/3)
```

## Box plot

Reference: [R plotly boxplots](https://plotly.com/r/box-plots/)

In plotly, we can use a loop to add each column's boxplot one at a time. In the 
following example, we turn off the legend, since the names also appear below 
each boxplot: 

```{r}
plt <- plot_ly(y = dt[, SysBP], type = "box", name = "SysBP",
    line = list(color = palette[1]), fillcolor = palette_a[1])  |> 
    add_boxplot(y = dt[, DiaBP], name = "DiaBP",
        line = list(color = palette[2]), fillcolor = palette_a[2])
plt |> layout(showlegend = FALSE)
```

## Grouped boxplot

```{r}
p <- plot_ly(y = dt[, Age], type = "box", color = dt[, Sex],
    colors = palette[2:1])
p |> layout(showlegend = FALSE)
```

## Histogram

```{r}
p <- plot_ly(
    x = dt[, Age], type = "histogram",
    nbinsx = 24,
    marker = list(color = palette_a[1], 
        line = list(color = palette[1], width = 2)
    )
) |> layout(xaxis = list(title = "Age (years)"))
p
```

## Grouped Histogram

```{r}
p <- plot_ly(x = dt[, Age], type = "histogram", color = dt[, Sex],
    colors = palette_a[2:1], nbinsx = 24)
p
```

More control is possible when adding traces one at a time:

```{r}
p <- plot_ly() |> 
    add_histogram(
        x = dt[Sex == "Female", Age],
        nbinsx = 24, name = "Female",
        marker = list(color = palette_a[2], 
        line = list(color = palette[2], width = 1))
    ) |> 
    add_histogram(
        x = dt[Sex == "Male", Age],
        nbinsx = 24, name = "Male",
        marker = list(color = palette_a[1], 
        line = list(color = palette[1], width = 1))
    ) |> 
    layout(xaxis = list(title = "Age"))
p
```

Can control placement of different traces' bars using `barmode` arguments, which defaults 
to "group".

To switch to "overlay":

```{r}
p |> layout(barmode = "overlay")
```

## Density plot

The built-in `density()` function returns x and y coordinates we can use to draw a
density plot.

```{r}
age_density <- density(dt[, Age])
```

```{r}
p <- plot_ly(
    x = age_density$x, 
    y = age_density$y, 
    type = "scatter", mode = "none",
    fill = "tozeroy",
    fillcolor = palette_a[1]
)
p
```

## Grouped density plot

```{r}
female_age_density <- density(dt[Sex == "Female", Age])
male_age_density <- density(dt[Sex == "Male", Age])
```

```{r}
p <- plot_ly(
    x = female_age_density$x, y = female_age_density$y,
    type = "scatter", mode = "none",
    name = "Female",
    fill = "tozeroy", fillcolor = palette_a[2]) |> 
    add_trace(
        x = male_age_density$x, y = male_age_density$y,
        type = "scatter", mode = "none",
        name = "Male",
        fill = "tozeroy", fillcolor = palette_a[1])
p
```

## Barplot

```{r}
schools <- data.frame(UCSF = 4, Stanford = 7, Penn = 12)
```

```{r}
schools_df <- data.frame(
    University = factor(colnames(schools), 
        levels = c("UCSF", "Stanford", "Penn")),
    N_schools = as.numeric(schools[1, ])
)
```

```{r}
p <- plot_ly(
    x = schools_df$University,
    y = schools_df$N_schools,
    name = "Schools",
    type = "bar",
    marker = list(color = palette_a[1], 
        line = list(color = palette[1], width = 2)
    ))
p |> layout(yaxis = list(title = "N schools"))
```

## Scatterplot

```{r}
p <- plot_ly(
    dt, x = ~Age, y = ~SysBP, 
    type = "scatter", mode = "markers",
    marker = list(color = palette_a[1])
)
p
```

same as:

```{r}
p <- plot_ly(dt, x = ~Age, y = ~SysBP) |> 
    add_trace(type = "scatter", mode = "markers",
        marker = list(color = palette_a[1])
    )
p
```

## Grouped Scatterplot

```{r}
p <- plot_ly() |> 
    add_trace(x = dt[Sex == "Female", Age], y = dt[Sex == "Female", SysBP],
        name = "Female",
        type = "scatter", mode = "markers",
        marker = list(color = palette_a[2])
    ) |> 
    add_trace(x = dt[Sex == "Male", Age], y = dt[Sex == "Male", SysBP],
        name = "Male",
        type = "scatter", mode = "markers",
        marker = list(color = palette_a[1])
    )
p
```

## Save plot to file

We'll use the grouped boxplot example from above to show how to save each type of plot
to file, using a PDF output as an example.

```{r}
#| eval: false
p <- plot_ly(y = dt[, Age], type = "box", color = dt[, Sex],
    colors = palette[2:1])
p |> layout(showlegend = FALSE)
save_image(p, "Age_by_Sex_plotly.pdf", width = 5.5, height = 5.5, scale = 1)
```

## References

- [R plotly Documentation](https://plotly.com/r/)
