# readr Data I/O {#sec-readrdataio}

## Read delimited data

The **readr** package includes functions `read_delim()`, `read_csv()`, `read_csv2()`, 
and `read_tsv()` to read delilmited files.

Important arguments:

- `delim`: The field delimiter, for `read_delim()` only. Defaults to `NULL`, which means
  it will try to guess the correct delimiter.
- `na`: Character vector of strings to interpret as missing values, i.e. `NA`.

```{r}
library(readr)
```

```{r}
dat <- read_csv("https://archive.ics.uci.edu/ml/machine-learning-databases/00519/heart_failure_clinical_records_dataset.csv")
```

As part of the tidyverse, **readr** returns a [tibble](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html).

```{r}
dat
```

## Write delimited data

The **readr** package includes functions `write_delim()`, `write_csv()`, `write_csv2()`.

```{r}
#| eval: false
write_csv(dat, "/path/to/file.csv")
```

## Further reading

- Read more on [tibbles](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html)
