---
execute:
  cache: true
knitr:
  opts_chunk: 
    comment: ''
---
# Vectorized Operations {#sec-vectorization}

Most built-in R functions are vectorized and many functions from external 
packages are as well.

```{block, type="rmdtip"}
A vectorized function operates on all elements of an object at the same time.
```

Vectorization is very efficient: it can save both human (your) time and machine 
time.

```{block, type="rmdtip"}
In many cases, applying a function on all elements simultaneously may seem 
obvious or expected behavior, but since not all functions are vectorized, make 
sure to check the documentation and/or test using a simple example.
```

## Operations between vectors of equal length

Such operations are applied between corresponding elements of each vector:

```{r}
x <- 1:10
z <- 11:20
x + z
```

i.e. the above is equal to `c(x[1] + z[1], x[2] + z[2], ..., x[n] + z[n])`

## Operations between a vector and a scalar

In this cases, the scalar is repeated to match the length of the vector, 
i.e. it is ***recycled***:

```{r}
x + 10
x * 2
x / 10
x ^ 2
```

## Operations between vectors of unequal length: **value recycling**

Operations between a vector and a scalar are a special case of operations 
between vectors of unequal length. Whenever you perform an operation between 
two objects of different length, 
***the shorter object's elements are recycled***:

```{r}
x + c(2:1)
```

```{block, type="rmdcaution"}
Operations between objects of unequal length can occur by mistake. If the 
shorter object's length is a multiple of the longer object's length, there will 
be ***no error or warning***, as above. Otherwise, 
there is a warning (which may be confusing at first) BUT 
***recycling still happens and is highly unlikely to be intentional***:
```

```{r}
x + c(1, 3, 9)
```

## Vectorized matrix operations

Operations between matrices are similarly vectorized, i.e. performed between 
corresponding elements:

```{r}
a <- matrix(1:4, 2)
b <- matrix(11:14, 2)
a
b
a + b
a * b
a / b
```

## Vectorized functions

Some examples of common mathematical operations that are vectorized:

```{r}
log(x)
sqrt(x)
sin(x)
cos(x)
```
